Last glacial cycle glacier erosion potential in the Alps

### 2021-05-25 articles/erosion/response-referee2.md

The manuscript text has been extended in several places. Please see our
responses to your specific comments below.

## Specific comments - Text:

> The 1 km resolution of the modelling is clearly important, so it is strange
> that the first mention is on p.11. Although it is very demanding of computer
> time, it is still (as admitted) not adequate to represent cirques (averaging
> of the order of 700 m across). Even stranger, there is no mention of the DEM
> used (it was SRTM in the 2018 paper). This information should be provided in
> section 2, p. 3-4.

The model horizontal resolutions of 1 and (for climate sensitivity runs) 2 km
is now mentioned in the methods. The input SRTM topography is also referenced
in both the methods and the caption of Fig.~2 where it is displayed in the
background. Following comments from reviewer #1, the discussion of model
limitations regarding the representation of cirque glaciers has also been
extended.

@@

> The lateral constraints on valley glaciers are usually covered by a ‘form
> factor’: how are they handled by this version of the PISM ice sheet model –
> how has lateral drag been included?

PISM uses a combination of shallow-ice and shallow-shelf stress balances,
neither of which includes lateral stresses. Glacier sliding velocities in
troughs are goverened by (the sliding law and) longitudinal stresses, which is
a limitation with respect to higher-order models. Following another comment
from reviewer #1 an additional paragraph on basal sliding uncertainties was
added, including the following sentence:

    Lateral stress gradients missing from the shallow-shelf approximation
    stress balance could also contribute to moderate sliding velocity in narrow
    troughs (Herman et al., 2011; Egholm et al., 2012a, b; Pedersen et al.,
    2014).

@@

> It might be pointed out that the ‘Koppes’ relationship with an exponent of
> 2.34 is based on 13 data points that are very poorly distributed: in fact,
> Koppes et al. (2015) go on to drop two outliers and reach an exponent of
> 2.62.

TODO: discuss Koppes without outliers.

> In lines 206-209, the point about trimlines, relating them to
> time-transgressive erosion, is well taken. But they represent the integrated
> effect of many phases of glacial erosion (as noted on 223-224), not
> necessarily just from recessional phases.

Indeed. The sentence was corrected to include "advancing and retreating
glaciers" instead of "retreating glaciers".

@@

## - Figures:

> In Fig. 3 is almost impossible to follow the brown lines, unless advancing
> and retreating parts of lines are distinguished (e.g. by colour). (The ice
> volume video may help out.)  Also: keep to ‘potential erosion volume’. The
> pale brown bars show a decline in annual erosion volume with ice volumes over
> 1.8 cm s.l.e., stronger than is implied in line 110 of the text – it is more
> than ‘slight’.

Figure 3 was reworked to use different colours for periods of increasing and
decreasing ice volume. This actually strikingly highlights the two different
regimes. Thank you for the suggestion. The label was changed to 'potential
annual erosion volume'. In the text, 'slight' was replaced with a 'general'
tendency for slower erosion during periods of extensive glaciation.

@@

> To aid interpretation of Figs 4 a-c and 7 a-d, a corresponding topographic
> map or hillshade of the Rhine area on the same scale should be provided.

In addition to the aforementioned changes to Fig. 4, a new panel was added for
the model final state at 0 ka, revealing the bedrock topography. So Fig. 4 now
has one panel during advance (36 ka), one during a maximum stage (24 ka), one
during retreat (16 ka), and one nearly ice-free (0 ka).

@@

> Fig. 1c illustrates a very unusual cirque, with a huge lake replacing a very
> unhealthy convex glacier. Fascinating, but surely a more representative
> cirque should be used. Actually these three photos are not well integrated
> with the text (pace lines 191 & 195-198). Lauterbrunnen is an extreme
> (vertical-sided) trough example, over-used by textbooks.

TODO: think about replacing the photos.

> Fig. 5: The elevation histogram in Fig. 5a could be misleading, as it seems
> to cover the whole study area, including parts never glacier-covered. It
> should be limited to areas glacier-covered at the maximum, and preferably
> supplemented by a histogram for those covered now or at the minimum. The
> accompanying main graph (5a) relates to areas glacier-covered at each time
> slice. Re another reviewer’s comment, yes LGM glaciers did reach below 500 m:
> at L. Garda (67 m?) and the Tagliamento Glacier lobe LGM ice reached below
> 100 m. An integrated plot of total-cycle potential erosion rate (and
> volume?) at each altitude, over glacier-covered areas, would be interesting
> (as an extra Figure): can rates at high altitude, where glacial duration is
> longer, catch up with those at low altitude?  Probably not, given the
> concentration of flow down major valleys. Striping above 3000 m, presumably
> related to low surface area, suggests a switch there to 20 m or 40 m bands
> would be advisable. Caption; ‘modelled potential erosion rates’?

The figure now includes a histogram of model domain elevations, a histogram of
glacier-covered elevations, and a plot of cumulative erosion potential volume
per elevation bands of 100 m. The latter shows that, as you suspected, high
altitude rapid erosion is largely offset by the limited number of grid cells
it concerns. Instead, the bulk of the erosion potential occurs below 2000 m.

TODO: hatch out regions with limited cell count?

@@

> In Fig. 7, the quantitative contrasts between the ‘Koppes’ law (e) and the
> others are alarming. If the scales are taken literally, (e) gives potential
> erosion around 1 m, while (f) and (g) give hundreds of m, in 120 ka. As 120
> ka is only a fraction of the Quaternary, hundreds of m seems too much, while
> 1 m seems very low (0.008 mm a-1 overall). For me, there should be a correct
> answer between these extremes. The spatial pattern, however, is more
> important.

Discussion on the magnitude of modelled erosion rates remains somewhat
speculative indeed due to the several sources of uncertainties, but the
relevant discussion text was extended and reworked:

    With a total Pleistocene glacial relief on the order of a kilometre
    (Preusser et al., 2011; Valla et al., 2011), a cumulative glacial erosion
    for the last glacial cycle in the order of 10-100 m can be expected.
    However, none of the tested erosion power-laws fall within this range.
    Instead, the erosion law calibrated on tidewater glaciers (Koppes et al.,
    2015) yields cumulative erosion in the Rhine Valley in the orders of
    metres, while the three erosion laws based on terrestrial glaciers
    (Humphrey and Raymond, 1994; Herman et al., 2015; Cook et al., 2020},
    result in kilometre-scale integrated erosion potential. During the Last
    Glacial Maximum and much of the last glacial cycle, Alpine paleoglaciers
    were closer in size, slope (an important parameter as we argue in the next
    section), and climatic context to the present-day glaciers of Patagonia and
    the Antarctic Peninsula (Koppes et al., 2015) than to Franz-Joseph Glacier
    (Herman et al., 2015) and many of the glaciers included in the global
    compilation by Cook et al., (2020). This may help to explain why the
    reality appears to fall in-between the tested erosion laws.

@@

> Fig. 8a shows a flat dome at 3000 m for about the first 20 km of the Rhine.
> As there are mountains at 3630, 3583 and 3192 m around the sources, it is
> likely that   there would be steeper ice slopes up to these peaks, giving
> rather different stresses and erosion potentials. Also are the instabilities
> in Fig. 8b, near source and terminus, edge effects or related to the 1 km
> resolution?

TODO: I think this is correct. Develop on basal shear stress patterns.

### 2021-05-26 articles/erosion/response-referee1.md

The field-based LGM outline was added on Fig. 2. Computing a percent overlap is
not straightforward. As for Fig. 4 in the 2018 paper, we use a manually
modified version of the Ehlers et al. (2011) data. The original data presents
numerous, expansive nunataks, many of which are found in accumulation area and
thus directly incompatible with the mass-balance scheme used in the 2018 paper.
However, the general eastward bias was discussed in two new sentences:

    It should be noted, however, that all runs presented here show a systematic
    bias with excessive glacier cover in the Eastern Alps and a diminished
    glacier extent in the Western Alps (Fig. 2a; further discussed in Seguinot
    et al., 2018). Thus the modelled patterns of erosion potential certainly
    includes a similar bias.

> Fig. 5 I have to confess Fig. 5a surprises me, there are really glaciers as
> low as 500 m.a.s.l (and even below) from 110 – 40 ka in the Seguinot et al.,
> 2018 model?

This is correct. It concerns only a small number of grid cells, primarily
located on the southern slope of the Alps where topography drops abruptly from
the highest peaks to deep valleys. This information was added to the plot by
hatching the regions with fewer than a hundred ice-covered grid cells. This
highlights that low-elevation values are not always representative, as can also
be confirmed in the new subplot showing the distribution of cumulative erosion
potential over 100 m elevation bins. The distribution of glacier cover per
elevation band over time (from which the hundred-cell contour was extracted)
will be included in a new version of the companion dataset.

### 2021-05-26 articles/erosion/response-referee2.md

PISM uses a combination of shallow-ice and shallow-shelf stress balances,
neither of which includes lateral stresses. Glacier sliding velocities in
troughs are goverened by (the sliding law and) longitudinal stresses, which is
a limitation with respect to higher-order models. Following another comment
from reviewer #1 an additional paragraph on basal sliding uncertainties was
added, including the following sentence:

    Lateral stress gradients missing from the shallow-shelf approximation
    stress balance could also contribute to moderate sliding velocity in narrow
    troughs (Herman et al., 2011; Egholm et al., 2012a, b; Pedersen et al.,
    2014).

> It might be pointed out that the ‘Koppes’ relationship with an exponent of
> 2.34 is based on 13 data points that are very poorly distributed: in fact,
> Koppes et al. (2015) go on to drop two outliers and reach an exponent of
> 2.62.

We clarified in the methods that we use  Koppes et al. (2015) erosion law
deriving from their full dataset. In addition, this sentence was added in the
discussion of the "choice of erosion law":

    On the other extreme, an even more non-linear erosion law, not tested here,
    derived from tidewater glaciers but excluding two outliers (ė = 5.3 × 10−9
    u_b^2.62, Koppes et al., 2015) would result in an even more localized
    pattern of erosion potential.

@@

> Fig. 1c illustrates a very unusual cirque, with a huge lake replacing a very
> unhealthy convex glacier. Fascinating, but surely a more representative
> cirque should be used. Actually these three photos are not well integrated
> with the text (pace lines 191 & 195-198). Lauterbrunnen is an extreme
> (vertical-sided) trough example, over-used by textbooks.

TODO: think about replacing the photos.

> Fig. 5: The elevation histogram in Fig. 5a could be misleading, as it seems
> to cover the whole study area, including parts never glacier-covered. It
> should be limited to areas glacier-covered at the maximum, and preferably
> supplemented by a histogram for those covered now or at the minimum. The
> accompanying main graph (5a) relates to areas glacier-covered at each time
> slice. Re another reviewer’s comment, yes LGM glaciers did reach below 500 m:
> at L. Garda (67 m?) and the Tagliamento Glacier lobe LGM ice reached below
> 100 m. An integrated plot of total-cycle potential erosion rate (and
> volume?) at each altitude, over glacier-covered areas, would be interesting
> (as an extra Figure): can rates at high altitude, where glacial duration is
> longer, catch up with those at low altitude?  Probably not, given the
> concentration of flow down major valleys. Striping above 3000 m, presumably
> related to low surface area, suggests a switch there to 20 m or 40 m bands
> would be advisable. Caption; ‘modelled potential erosion rates’?

The figure now includes a histogram of model domain elevations, a histogram of
glacier-covered elevations, and a plot of cumulative erosion potential volume
per elevation bands of 100 m. The latter shows that, as you suspected, high
altitude rapid erosion is largely offset by the limited number of grid cells
it concerns. Instead, the bulk of the erosion potential occurs below 2000 m.
To highlight periods when low-elevation is relevant (and address the comment
from reviewer #1), we also hatched elevation bands that contain fewer than a
hundred ice-covered grid cells for a given time.

@@

Discussion on the magnitude of modelled erosion rates remains somewhat
speculative indeed due to the several sources of uncertainties, but the
relevant discussion text was extended and reworked:

    With a total Pleistocene glacial relief on the order of a kilometre
    (Preusser et al., 2011; Valla et al., 2011), a cumulative glacial erosion
    for the last glacial cycle in the order of 10-100 m can be expected.
    However, none of the tested erosion power-laws fall within this range.
    Instead, the erosion law calibrated on tidewater glaciers (Koppes et al.,
    2015) yields cumulative erosion in the Rhine Valley in the orders of
    metres, while the three erosion laws based on terrestrial glaciers
    (Humphrey and Raymond, 1994; Herman et al., 2015; Cook et al., 2020},
    result in kilometre-scale integrated erosion potential. During the Last
    Glacial Maximum and much of the last glacial cycle, Alpine paleoglaciers
    were closer in size, slope (an important parameter as we argue in the next
    section), and climatic context to the present-day glaciers of Patagonia and
    the Antarctic Peninsula (Koppes et al., 2015) than to Franz-Joseph Glacier
    (Herman et al., 2015) and many of the glaciers included in the global
    compilation by Cook et al., (2020). This may help to explain why the
    reality appears to fall in-between the tested erosion laws.

> Fig. 8a shows a flat dome at 3000 m for about the first 20 km of the Rhine.
> As there are mountains at 3630, 3583 and 3192 m around the sources, it is
> likely that   there would be steeper ice slopes up to these peaks, giving
> rather different stresses and erosion potentials. Also are the instabilities
> in Fig. 8b, near source and terminus, edge effects or related to the 1 km
> resolution?

This is correct. The "basal drag" quantity plotted on Fig. 8b is the magnitude
of a two-component vector: the basal shear stress. The basal shear stress
itself strongly depends on surface altitude gradients, but this includes both
along-flow slopes visible on the topographic profile of Fig. 8a, and the
invisible across-flow slopes. While the upper part of the topographic profile
appears smooth, there are indeed steep (bedrock and surface) slopes on either
side of the profile in this region. Because the upper part of the valley is
narrow, the glacier center-line fluctuating over time, and the plot
interpolated between 1-km grid cells, these lateral components enter the
computation of the basal shear stress magnitude in the higher reaches.

Second, the activation of sliding also increases basal drag, which explains
that basal drag also fluctuates together with the yield stress. This is
particularly affecting the lower part of the profiles. Finally, normalization
by overburden pressure amplifies all stress variations where ice thickness is
small. But due to changes in glacier thickness, it would not be possible to
compare different profiles without such normalization.

While we would rather not include such details in the paper, we tried our best
to summarise the effect with this additional sentence in the figure caption:

    Some of the observed basal drag fluctuations in the upper part of the
    transect result from steep slopes on either side of the narrow valley.

### 2021-05-26 articles/erosion/alps-erosion.tex
Update hypsogram caption and responses.
Discuss basal drag fluctuations in upper Rhine.

    \begin{figure*}
      \centerline{\includegraphics{alpero_hypsogram}}
      \caption{%
        \textbf{(a)} Potential erosion rate ``hypsogram'', showing the
          geometric mean of (non-zero) modelled rates in 10-m elevation bands
          across the entire model domain and its time evolution.
        \textbf{(b)} Distribution of model domain bedrock topography (grey
          bars) and glaciated topography (blue bars) in 100-m elevation bands,
          and the corresponding cumulative potential erosion volume (brown
          line).
        \textbf{(c)} Same as Fig.~\ref{fig:cumulative}b.
        \todo{fix labels, test hatching below a minimum number of grid cells}}
      \label{fig:hypsogram}
    \end{figure*}

@@

    That said, the magnitude of the modelled erosion rates can hardly be used
    as a criteria to rule out this or that erosion law for the glacial Alps.
    While sliding is here modelled by a pseudo-plastic law based on
    till dilatation under high water pressure \citep{Tulaczyk.etal.2000},
    empirical erosion rules are thought to be a proxy for glacier abrasion
    (cf.~\citealp{Hallet.1979}; and possibly plucking) of a glacier flowing
    over a hard bed. Current erosion rates in the Alps strongly vary depending
    on the substratum \citep{Steinemann.etal.2020, Steinemann.etal.2021}, and
    it is arguable that a layer of till and proglacial sediments may have
    armored the glacier bed from erosion for at least parts of the glacial
    cycle. Thus, given the lack of data to constrain the erosion rule, it can
    not be excluded that a linear or sub-linear erosion law
    \citep[e.g.,][]{Cook.etal.2020} would lead to a less localized pattern of
    erosion (Fig.~\ref{fig:powerlaws}c and~d), and a stronger correlation
    between modelled ice volume and erosion (Fig.~\ref{fig:powerlaws}k and~l).
    \todo{discuss Koppes without outliers}

@@

    \begin{figure}
      \centerline{\includegraphics{alpero_basaldrag}}
      \caption{%
        \textbf{(a)} Modelled basal and surface topographies interpolated along
        the Rhine Glacier during glacier advance (36\,ka), maximum extent
        (24\,ka) and retreat (16\,ka, using the same transect as in
        Fig.~\ref{fig:basaldrag}).
        \textbf{(b)} Interpolated yield stress in the pseudo-plastic sliding law
        (dotted lines) and magnitude of the basal shear stress (basal drag,
        solid lines), both normalized against the ice overburden pressure.
        Basal drag values greater than yield stress result in glacier sliding
        \citep[cf.][for model physics]{Seguinot.2014, Seguinot.etal.2016}.}
      \label{fig:basaldrag}
    \end{figure}

>>>>

    \begin{figure*}
      \centerline{\includegraphics{alpero_hypsogram}}
      \caption{%
        \textbf{(a)} Potential erosion rate ``hypsogram'', showing the
          geometric mean of (non-zero) modelled rates in 10-m elevation bands
          across the entire model domain and its time evolution. Hatches
          indicate elevation bands with fewer than a hundred grid cells
          (100\,$km^2$).
        \textbf{(b)} Distribution of model domain bedrock topography (grey
          bars) and glaciated topography (blue bars) in 100-m elevation bands,
          and the corresponding cumulative potential erosion volume (brown
          line).
        \textbf{(c)} Same as Fig.~\ref{fig:cumulative}b.
        \todo{fix labels, test hatching below a minimum number of grid cells}}
      \label{fig:hypsogram}
    \end{figure*}

@@

    That said, the magnitude of the modelled erosion rates can hardly be used
    as a criteria to rule out this or that erosion law for the glacial Alps.
    While sliding is here modelled by a pseudo-plastic law based on
    till dilatation under high water pressure \citep{Tulaczyk.etal.2000},
    empirical erosion rules are thought to be a proxy for glacier abrasion
    (cf.~\citealp{Hallet.1979}; and possibly plucking) of a glacier flowing
    over a hard bed. Current erosion rates in the Alps strongly vary depending
    on the substratum \citep{Steinemann.etal.2020, Steinemann.etal.2021}, and
    it is arguable that a layer of till and proglacial sediments may have
    armored the glacier bed from erosion for at least parts of the glacial
    cycle. Thus, given the lack of data to constrain the erosion rule, it can
    not be excluded that a linear or sub-linear erosion law
    \citep[e.g.,][]{Cook.etal.2020} would lead to a less localized pattern of
    erosion (Fig.~\ref{fig:powerlaws}c and~d), and a stronger correlation
    between modelled ice volume and erosion (Fig.~\ref{fig:powerlaws}k and~l).
    On the other extreme, an even more non-linear erosion law, not tested here,
    derived from tidewater glaciers but excluding two outliers
    \citep[${\dot{e}=5.3\times10^{-9}u_\mathrm{b}^{2.62}}$,][]{Koppes.etal.2015}
    would result in an even more localized pattern of erosion potential.

@@

    \begin{figure}
      \centerline{\includegraphics{alpero_basaldrag}}
      \caption{%
        \textbf{(a)} Modelled basal and surface topographies interpolated along
        the Rhine Glacier during glacier advance (36\,ka), maximum extent
        (24\,ka) and retreat (16\,ka, using the same transect as in
        Fig.~\ref{fig:basaldrag}).
        \textbf{(b)} Interpolated yield stress in the pseudo-plastic sliding law
        (dotted lines) and magnitude of the basal shear stress (basal drag,
        solid lines), both normalized against the ice overburden pressure.
        Basal drag values greater than yield stress result in glacier sliding
        \citep[cf.][for model physics]{Seguinot.2014, Seguinot.etal.2016}.
        Some of the observed basal drag fluctuations in the upper part of the
        transect result from steep slopes on either side of the narrow valley.}
      \label{fig:basaldrag}
    \end{figure}
